---
title: 'Reproducible Research: Analyzing activity patterns'
author: "Matthias Schmitt"
date: "14 Januar 2017"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=10, fig.height=6)

library(lubridate)
library(ggplot2)
library(plyr)
```

## Project goal & data setup  

As described on the course website, we use ''data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.'' We analyze a set of questions related to activity patterns in the data, especially the difference between weekdays and weekends. We first read the data:


```{r}
# for local data file
# activity <- read.csv("D:/Documents/R/data-science/reproducible/activity.csv")

# auto-download from web source
temp <- tempfile()
download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",temp)
unzip(temp)
unlink(temp)

activity <- read.csv("activity.csv")
```

Since we need to work with the *date* variable a lot later, we transform it here, and introduce the flag *isWeekend*, which we need later to analyse patterns on weekends separately from those during the week. 

```{r}
activity$date <- as.Date(activity$date)

# there's a base R function for getting the weekdays, but I use the lubridate version
# to make it work on all locales (I have a non-US locale windows OS)
activity$wday <- wday(activity$date, label=TRUE)
activity$isWeekend <- activity$wday %in% c("Sat", "Sun")

```

## Total number of steps taken each day

We aggregate the number of steps for each date in the data set, and distinguish between weekend and weekday for each day:

```{r}
steps_per_day <- aggregate(steps ~ date+isWeekend, activity, sum)
```

Let's explore this as a time series - this was not requested, but it's informative:

```{r}
ggplot(steps_per_day, aes(x=date, y=steps, fill=isWeekend)) + geom_bar(stat="identity") + labs(x="Date", y="# of steps") + scale_x_date()
```

For some days there seems to be no activity at all, which is quite implausible. We're therefore tasked
to impute activity data for those dates. We'll address that below, first we plot the requested histogram of steps taken per day:

```{r}
ggplot(steps_per_day, aes(steps, fill=isWeekend)) + geom_histogram(bins = 20) + labs(x="# of steps", y="# of days")
```

## Mean and median number of steps taken each day

```{r}
# use summary table 
summary(steps_per_day)
# ... or alternatively we could use mean() and median() and produce a table using xtable
```

The mean is 10766 steps per day, the median is 10765 steps per day.

## Time series of the average number of steps taken

We now focus on the within-day pattern (intervals during the day) and average over all dates in the dataset.

```{r}
steps_per_interval <- aggregate(steps ~ interval, activity, mean)
ggplot(steps_per_interval, aes(x=interval, y=steps)) + geom_line()+ labs(x="time interval", y="# of steps")
```

The 5-minute interval that, on average, contains the maximum number of steps is

```{r}
highest_activity <- steps_per_interval[which.max(steps_per_interval$steps),]
print(highest_activity)
```

## Imputing missing data 

Above we saw that on some dates no steps were measured at all. Whatever the reason (maybe the owner forgot to switch the device on), we're tasked to impute data for the dates were no steps were measured.
Because activity patterns differ between weekend and weekday (shown below in detail), we're careful to impute a separate data point for each day of the week separately. We compute the mean of all the 5min interval for each day of the week and aggregate.
 
```{r}
steps_per_wday <- aggregate(steps ~ interval+wday, activity, mean)
steps_per_wday <- rename(steps_per_wday, c("steps"="imputed_steps"))

head(steps_per_wday)

activity_imputed <- merge(activity, steps_per_wday, all.x=TRUE)
activity_imputed$steps <- ifelse(is.na(activity_imputed$steps), activity_imputed$imputed_steps, activity_imputed$steps)
activity_imputed$isImputed <- activity_imputed$imputed_steps == activity_imputed$steps

steps_per_day.imputed <- aggregate(steps ~ date+isImputed, activity_imputed, sum)
```

The time series now looks as follows:

```{r}
ggplot(steps_per_day.imputed, aes(x=date, y=steps, fill=isImputed)) + geom_bar(stat="identity") + labs(x="Date", y="# of steps") + scale_x_date()
```

which is quite plausible. The histogram of the total number of steps taken each day after missing values are imputed is as follows:

```{r}
ggplot(steps_per_day.imputed, aes(steps, fill=isImputed)) + geom_histogram(breaks=seq(1, 24000, by = 2000)) + labs(x="# of steps", y="# of days")
```

Notice that we start the binning of the histogram at 1. That's to account for the fact that for most dates we didn't impute any data, and the count of those dates (roughly 40) for show up at $x = 0$.

## Comparison of activity pattern on weekdays versus weekend 

Last but not least we're tasked to compare the average number of steps taken per 5-minute interval across weekdays and weekends in a time series plot:

```{r, echo=TRUE}
steps_per_interval <- aggregate(steps ~ interval+isWeekend, activity, mean)
ggplot(steps_per_interval, aes(x=interval, y=steps)) + geom_line() + labs(x="time interval", y="# of steps") + facet_grid(isWeekend ~ .,  labeller = as_labeller(c('FALSE'="Weekday", 'TRUE'="Weekend"))) 
```

On weekdays there's a clear peak in activity around interval 800, after which activity is reduced. On weekends activity is more distributed, without a clear peak. Overall, activity is higher on weekends than on weekdays.

